Hydra battles in Coq

### 2020-09-16 doc/chapter-ordinals.tex
minor modifications

@@

Finally the nature of any ordinal (inside this notation system) is decidable:

\begin{Coqsrc}
Lemma ZLS_dec (alpha : t) :
  {alpha = 0} +
  {Limit alpha} +
  {beta : t | Successor alpha beta}.
\end{Coqsrc}

@@

@@

>>>>

\subsection{Ordinal notations and measures for proving termination}
\label{sect:measure-ON}

The following lemma (together with the type class mechnism) allows us to use simply  measures towards an ordinal notation. It is just an application of  the libraries \texttt{Coq.Wellfounded.Inverse\_Image}
and  \texttt{Coq.Wellfounded.Inclusion}.

\begin{Coqsrc}
Definition m_lt {A:Type}{lt: relation A}
            {compare : A -> A -> comparison}
            {on : ON lt compare}
            {B : Type}
  (m : B -> A) : relation B :=
  fun x y =>  on_lt (m x) (m y).

Lemma wf_measure  {A:Type}(lt: relation A)
            {compare : A -> A -> comparison}
            {on : ON lt compare}
            {B : Type}
            (m : B -> A) :
  well_founded (m_lt m). 
\end{Coqsrc}

A simple example of application is given in Sect.~\vref{sect:merge-example}.

@@

Finally, the nature of any ordinal is decidable(inside this notation system) :

\noindent\emph{From Module~\href{../src/html/hydras.OrdinalNotations.Generic.html}{OrdinalNotations.Generic}}
\begin{Coqsrc}
  Definition ZeroLimitSucc_dec {A:Type}{lt: relation A}
           {compare : A -> A -> comparison}
           {on : ON lt compare} :=
  forall alpha,
    {Least alpha} +
    {Limit alpha} +
    {beta: A | Successor alpha beta}.
\end{Coqsrc}

\begin{Coqsrc}
Definition Zero_limit_succ_dec : ZeroLimitSucc_dec.
\end{Coqsrc}

@@

\subsection{Arithmetic of \texorpdfstring{$\omega^2$}{omega^2}}

\subsubsection{Addition}

We can define on \texttt{ON\_omega2} an addition which extends the addition on \texttt{nat}.

\begin{Coqsrc}
Definition  plus (alpha beta : t) : t :=
  match alpha,beta with
  | (0, b), (0, b') => (0, b + b')
  | (0,0), y  => y
  | x, (0,0)  => x
  | (0, b), (S n', b') => (S n', b')
  | (S n, b), (S n', b') => (S n + S n', b')
  | (S n, b), (0, b') => (S n, b + b')
   end.

Infix "+" := plus : o2_scope.
\end{Coqsrc}

Please note that this operation is not commutative:

\begin{Coqsrc}
Example non_commutativity_of_plus :  omega + 3 <> 3 + omega.
Proof.
  cbn.
\end{Coqsrc}

\begin{Coqanswer}
1 subgoal (ID 237)

  ============================
(1, 3) <> omega.
\end{Coqanswer}

\begin{Coqsrc}
 discriminate.
Qed.
\end{Coqsrc}

\subsubsection{Multiplication}

The restriction of ordinal multiplication to the segment $[0,\omega^2)$ is not a total function.
For instance $\omega\times\omega= \omega^2$ is outside the set of represented values.
Nevertheless, we can define two operations mixing natural numbers and ordinals.

\begin{Coqsrc}
(** multiplication of an ordinal and a natural number *)

Definition mult_fin  (alpha : t) (p : nat): t :=
  match alpha, p with
 |  (0,0), _  => zero
 |  _, 0 => zero
 |  (0, n), p => (0, n * p)
 |  ( n, b),  n' => ( n *  n', b)
               end.
Infix "*" := mult_fin : o2_scope.

(** multiplication of  a natural number and an ordinal *)

Definition fin_mult (n:nat)(beta : t) : t :=
  match n, beta with
 |  0, _  => zero
 |  _, (0,0) => zero
 |   n , (0,n') => (0, (n*n')%nat)
 |  n, (n',p') => (n', (n * p')%nat)
 end.

Example e1 : fin_mult 3 (omega * 7 + 15) = omega * 7 + 45.
Proof. reflexivity. Qed.

Example e2 :  omega * 2 + 2 + omega * 3 + 10 = omega * 5 + 10.
Proof. reflexivity. Qed.
\end{Coqsrc}

Multiplication with a finite ordinal and addition are related through the following lemma:

\begin{Coqsrc}
Lemma unique_decomposition alpha : 
    exists! i j: nat,  alpha = omega * i + j.
\end{Coqsrc}

\subsection{A proof a termination using \texorpdfstring{$\omega^2$}{omega^2}}

Using the lemma of Sect.~\vref{sect:measure-ON}, we can define easily the function which merges two

\begin{Coqsrc}
(* adapted from Pascal Manoury et al. *)

Require Import Coq.Program.Wf.

Section Merge.

  Variable A: Type.

  Local Definition m (p : list A * list A) :=
    omega * length (fst p) + length (snd p).

  Program Fixpoint  merge  (ltb: A -> A -> bool)
          (xys: list A * list A)
          {wf (m_lt m) xys} :
    list A :=
    match xys with
      (nil, ys) => ys
    | (xs, nil) => xs
    | (x :: xs, y :: ys) =>
      if ltb x y then x :: merge  ltb (xs, (y :: ys))
      else y :: merge  ltb ((x :: xs), ys)
    end.

  Next Obligation.
    unfold m, m_lt; destruct xs; simpl; left; lia.
  Defined.

  Next Obligation.
    unfold m, m_lt ;cbn; destruct ys; simpl; right; lia.
  Defined.

End Merge.
\end{Coqsrc}

@@

\index{Projects}
\begin{project}
Prove that \texttt{Omega\_plus\_Omega} cannot be a sub-notation of \texttt{Omega}.
\end{project}
